---
title: "Overview of the CTNote Library"
author: "Gabriel Odom, Laura Brandt, Sean Luo, and Ray Balise"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{CTNote Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, load-packages, include=FALSE}
library(ctn0094data)
library(ctn0094DataExtra)
library(CTNote)
library(readxl)
library(kableExtra)
library(tidyverse)
```




# Introduction and Motivation


## Background

Over 750,000 Americans have died from a drug overdose [since 1990](https://www.hhs.gov/opioids/about-the-epidemic/opioid-crisis-statistics/index.html). [Since 2018](https://www.pewtrusts.org/en/research-and-analysis/fact-sheets/2020/12/medications-for-opioid-use-disorder-improve-patient-outcomes), 2 in 3 drug overdose deaths have been from an opioid (almost 50,000 deaths). Nearly 1.3 million Americans are currently using one or more of the following medications to treat opioid use disorder (OUD): methadone, buprenorphine, and naltrexone. The [National Institute on Drug Abuse](https://nida.nih.gov/)'s [Clinical Trials Network](https://nida.nih.gov/about-nida/organization/cctn/clinical-trials-network-ctn) have funded over 100 clinical trials and clinical trial supplemental studies to study the effects of new and existing medications to treat addiction. Furthermore, opioid use disorder has in the past been predominately a disease of [middle-class white Americans](http://www.doi.org/10.1007/s11013-016-9496-5). However, these demographics have been shifting, and with this shift come exacerbated [health disparities](https://www.addictioncenter.com/news/2019/10/racial-disparities-opioid-addiction-treatment/). 

Clinical trials assess the effectiveness of medication-based treatments for OUD (MOUD). A wide variety of treatment outcomes have been used in clinical trials since 1972 (Table 1 shows outcomes from 41 papers assessing MOUD). These endpoints all assess participant success or failure in treatment via the results from scheduled *urine drug screenings* (UDS) or *urine opioid screenings* (UOS). Accurately measuring the efficacy of MOUD for clinical trial participants is paramount, but no "gold standard" trial outcomes exist.

Moreover, trial outcomes commonly used in CTN trials have not been widely assessed for racial / ethnic measurement variance, but some such effects have already been described. For example, racial/ethnic bias was assessed for [OUD trial completion rates](https://www.doi.org/10.1016/j.drugalcdep.2018.06.006), and this treatment outcome was found to change disproportionately based on participant race/ethnicity in large US cities. In clinical trials for OUD, if a treatment outcome has latent racial/ethnic bias (after controlling for relevant clinical factors), then one of two scenarios will occur: (1) if minority subjects are disproportionately marked as treatment successes in error, then the observed quality of an otherwise ineffective therapeutic will be inflated for that subgroup; conversely, (2) if minority subjects are disproportionately marked as treatment failures in error, then the observed quality of an otherwise effective experimental therapeutic will be diminished. Treatment outcomes which depend on race/ethnicity will lead to inequitable over/under-treatment of minority individuals who suffer from OUD.

```{r, import-table-one, include=FALSE}
pathToTable1_char <- system.file(
  "suppl_docs", "definitions_20220405.xlsx", package = "CTNote", mustWork = TRUE
)
tab1_df <- readxl::read_xlsx(pathToTable1_char) 
```

```{r, show-table-one, results='asis'}
tab1_df %>% 
  select(-`Frequency of UOS`, -`Coded column name`, -DOI) %>% 
  rename(
    Group = `Outcome Group`,
    Endpoint = `Primary Endpoint`,
    Class = `Numeric Class`,
    Definition = `Definition/Assessment of Outcome`,
    `Missing is` = `Missing UOS coded as`
  ) %>% 
  mutate(
    Group = case_when(
      str_detect(Group, "Abstinence") ~ "Abstinence",
      str_detect(Group, "Relapse") ~ "Relapse",
      str_detect(Group, "Reduction") ~ "Reduction",
    )
  ) %>% 
  kable("html") %>%
  column_spec(1:4, width = "3cm") %>%
  column_spec(5, width = "5cm") %>%
  kable_styling("striped", font_size = 11) %>%
  kable_minimal() %>% 
  # All styling and spec calls have to come BEFORE this line.
  scroll_box(width = "1000px", height = "500px") 
```

Each row of Table 1 is an outcome used in a past clinical trial for MOUD. As preliminary work, our clinical team grouped these by clinical practice into three classes: *abstinence*, *relapse*, and substance use *reduction*. These classes were further split into subclasses based on their clinical interpretation. Of note, these metrics also differed in how they considered a scheduled but incomplete UDS (i.e. a "missing" UDS). We removed outcomes which could not be used to measure subject-specific outcomes (e.g. "proportion of subjects in treatment group with positive UDS in study week 3"); we removed outcomes which could not yield a univariate metric of subject success (e.g. "number of positive UDS per three-week window over time"); we split papers which assessed multiple outcomes (e.g. complete abstinence between weeks 5-12 AND total number of abstinent weeks), and we combined papers which used identical outcomes. At the end of our literature search, we have `r nrow(tab1_df)` outcomes from registered clinical trials to consider.


## Objectives

> Clinically meaningful, patient-centric endpoints beyond abstinence are needed to define success in clinical trials. - Dr. Nora Vulkow, Address to CTN, April 2022

This is the first large-scale, empirical (data driven) comparison of MOUD clinical trial endpoints, and to our knowledge, this is the first racial/ethnic sensitivity analysis of multiple clinical trial outcome definitions. Furthermore, this is the only standard and code-based library of treatment outcome definitions to date, and we believe it will be useful to all future substance use disorder clinical trials in NIDA's [Clinical Trials Network](https://nida.nih.gov/about-nida/organization/cctn/clinical-trials-network-ctn) (CTN) and in other substance use disorder research.

Our scientific questions are as follows: (1) When applied to the same clinical trials data, do these outcomes follow the same clinically intuitive clusters (abstinence, relapse, use reduction)? (2) When applied to the same clinical trials data, *and* controlling for clinically relevant confounders, are some of these outcomes more sensitive to race/ethnicity than others?


*******************************************************************************
</br>


# Methods


## The Clinical Trials Data

The [CTN-0094 Project](http://ctndisseminationlibrary.org/protocols/ctn0094.htm) research team harmonized data from 3 large-scale MOUD clinical trials: [CTN-0027](http://ctndisseminationlibrary.org/protocols/ctn0027.htm): "Starting reatment with Agonist Replacement Therapies (START)"; [CTN-0030](http://ctndisseminationlibrary.org/protocols/ctn0030.htm): "Prescription Opiate Abuse Treatment Study (POATS)"; and [CTN-0051](http://ctndisseminationlibrary.org/protocols/ctn0051.htm): "Extended-Release Naltrexone vs. Buprenorphine for Opioid Treatment (X:BOT)". These three are nationally representative, prospective clinical trials with over 3600 combined participants. These trials treated participants for 16-24 weeks, and collected rich baseline substance use data, demographics, and medical/psychiatric evaluations. 

We have `r nrow(derived_weeklyOpioidPattern)` fully de-identified subjects with demographics and weekly UDS results for the months they were in treatment, 2492 of whom completed randomization to a treatment arm. This harmonized data will be released in the R packages `ctn0094data` and [`ctn0094DataExtra`](https://github.com/CTN-0094/ctn0094DataExtra). To simplify our analysis of endpoint sensitivity to race/ethnicity, we coded race and ethnicity conjointly as "Non-Hispanic White" ($n_W = 2484$), "Non-Hispanic Black" ($n_B = 347$), "Hispanic" ($n_H = 507$), and "Other" ($n_O = 222$). In order to summarize complex substance use patterns over the weeks of the multiple clinical trials, we standardized all dates to the day of trial consent, and we coded the urinalysis results for each day/week as a "word".


## Substance Use Pattern "Word"

As described in detail in [Odom et al, (2022)](<ctn0094DataExtra Vignette / Publication here>), we create a [sufficient statistic](https://doi.org/10.1098/rsta.1922.0009) of all weekly UDS results. For a substance of interest (opioids in our case), this statistic is a compact representation of the full pattern of substance use for an individual participant. To be of use, this summary statistic must have the following properties:

1. It can be directly parsed by a computer
2. It can be directly interpreted by a human
3. It represents all of the same information about participant use of a substance or group of substances that would be present in a medical summary of the participant


### Pattern Definition
In order to compactly summarize a participant's pattern of use for a particular substance over time, we first define the following legend:

- **+**: positive for the substance(s) in a specified window of time (a day, week, month, etc.) by urine screen (or self report, if such data are of interest)
- **–**: negative for the substance(s)
- **o**: subject failed to provide a urine sample
- <b>_</b>: no specimens required (weekends, holidays, pre-randomization period, alternating visit days/weeks)
- <b>*</b>: inconclusive results or mixed results (e.g. subject provided more than one urine sample in the time interval, and they did not agree)


### Use Pattern Examples
```{r}
egWho_int <- 2089L
```

We will first observe the recorded opioid use data via urine screen for subject `r egWho_int` after randomization.
```{r results='asis'}
ctn0094data::all_drugs %>% 
  filter(who == egWho_int) %>% 
  filter(source == "UDSAB") %>% 
  filter(when >= 0) %>% 
  arrange(when) %>% 
  kable() %>% 
  scroll_box(width = "500px", height = "500px") 
```

While we observe that this participant used various substances throughout the clinical trial, it's difficult to make clinical judgement using the data in this form. On the one hand, this data can be clearly parsed by a computer, and it represents all of the opioid use for this subject. However, data in this form is not easy to directly interpret by a clinician.

Now observe the opioid use pattern summary for this participant: 
```{r}
egUsePattern_char <- 
  ctn0094DataExtra::derived_weeklyOpioidPattern %>% 
  filter(who == egWho_int) %>% 
  # select(who, `Weeks of Treatment` = endWeek, `Opioid Use Pattern` = Phase_1)
  pull(Phase_1)
egUsePattern_char
```

Using the basic pattern definition above, we can clearly see that this subject had a challenging first month (`++++`), improved in the second month (`---+`), and then remained abstinent from opioids for the remainder of the clinical trial (`--------------o-`). Participant substance use pattern data in this form *is* easy to directly interpret by a clinician. Moreover, this symbolic representation of substance use can be parsed and summarized by a computer. Finally, notice that this pattern represents the entire course of treatment for subject `r egWho_int` without loss of any clinically relevant information about their opioid addiction/dependence.


### Use Pattern Limitations    
While this substance use pattern summary satisfies the three conditions necessary to be a "sufficient" statistic mentioned previously, there are three main limitations:

The first limitation surrounds the idea of poly-drug use. These use pattern summaries are substance or substance group specific. Notice that, while this participant does appear to curb their weekly opioid and heroin use during the course of treatment, their cocaine use does *not* decrease over the same time interval. However, the **opioid** use pattern summary can not display information about concurrent **cocaine** use. While this is a current limitation, one of our current areas of research involves expanding the support of substance use pattern "words" to include poly-substance use in a meaningful way (including thee ability to preserve cross-substance correlations).

The second limitation is one of parsimony. In order to ensure that use pattern summaries work regardless of a computer's locale or operating system, we limit the symbols in the "word" to proper symbols from the [American Standard Code for Information Interchange](https://www.ascii-code.com/) (ASCII) list. Technically speaking, there are 128 [7-bit ASCII](https://www.ibm.com/docs/en/aix/7.1?topic=support-ascii-characters) symbols, but only 96 are visible (printable) characters. Put simply, these 96 characters are all the symbols that a traditional North American computer keyboard can make. Therefore, we could define a substance use pattern "word" with any combination of these 96 printable symbols. However, such a visual summary would be incredibly challenging to interpret, consequently failing to meet the requirement that our summary be easily read by a clinician. In the interest of simplicity, we chose the five symbols mentioned above (`+`, `-`, `o`, `*`, and `_`). We recognize that different clinical trial designs may necessitate the introduction of additional symbols to this use pattern summary, but we strongly recommend (due to human [memory constraints](https://doi.org/10.1037/h0043158)) that such lists be kept to seven symbols or fewer. As an example, we discussed the benefit of adding a special symbol for a "missing but excused" clinic visit, but we ultimately did not because such instances were rare in our data.

The third limitation is due to the cold and unfeeling logic of a computer. The computer cannot understand the many extenuating circumstances that pervade our human experience. Worth repeating is the old maxim of computing: "computers give you what you ask for, not what you want." Consider a concrete example: assume a participant was supposed to visit the clinic for a weekly urine test on Thursday, but they came on Wednesday instead due to a conflict at their work; their urine sample from Wednesday came back "clean". What would a nurse at the substance use clinic do in this case? Most people are reasonable; they would count the urine screen for that week as negative for opioids. What would a computer do? Without the direct input of a human to override the clinical trial protocol, a computer would count the negative urine on Wednesday but still mark the participant as a "failure to appear" / "missing urine screen" on Thursday. In many clinical trial protocols, missing urine screens are imputed as "positive" for the substance of interest, so now this participant's use pattern summary is marked `*` (mixed results) instead of `-` (negative results) for that week.


## Analyzing Use Patterns with `CTNote`
- The `CTNote::` Package Helper Functions
    + Handling missing data: `recode_missing_visits()` and `impute_missing_visits()`
    + Accounting for the study observation design: `collapse_lattice()` and `view_by_lattice()`
    + Detecting a use pattern: `detect_subpattern()` and `detect_in_window()`
    + Measuring lengths of consecutive behavior: `measure_retention()` and `measure_abstinence_period()`
    + Counting substance use events: `count_matches()`


*******************************************************************************
</br>


# Results

- Examples of Treatment Outcome Definitions
- Clustering Treatment Outcomes
    + Resampled Dendrogram (Figure 1)
- Assessing Racial/Ethnic Sensitivity in Outcomes


*******************************************************************************
</br>


# Discussion
